---
title: "Homework 5: Matrices, lists and data frames"
author: "Aura M. Alonso-Rodríguez"
date: "February 12, 2020"
output: html_document
---

<br />
Full instructions for this homework are available [here](https://gotellilab.github.io/Bio381/Homeworks/Homework05_S2020.html). 

<br />

##### Exercise 1  
<br />

+ Assign to the variable n_dims a single random integer between 3 and 10.  
```{r}
n_dims <- sample(3:10, 1)
print(n_dims)
```

+ Create a vector of consecutive integers from 1 to n_dims2.
```{r}
my_vec <- c(1:(n_dims^2))
print(my_vec)
```

+ Use the sample function to randomly reshuffle these values.
```{r}
my_vec2 <- sample(my_vec)
print(my_vec2)
```

+ create a square matrix with these elements.
+ print out the matrix.
```{r}
my_mat <- matrix(data=my_vec2, nrow=n_dims, ncol=n_dims)
print(my_mat)
```


+ find a function in r to transpose the matrix.  
  print it out again and note how it has changed.
```{r}
my_mat2 <- t(my_mat)
print(my_mat2)
```

+ calculate the sum and the mean of the elements in the first row and the last row.
```{r}
sum(my_mat2[1,])
sum(my_mat2[n_dims,])
mean(my_mat2[1,])
mean(my_mat2[n_dims,])

sum(my_mat2[1,], my_mat2[n_dims,])
mean(c(my_mat2[1,],my_mat2[n_dims,]))

```

+ read about the eigen() function and use it on your matrix
```{r}
z <- eigen(my_mat2)
print(z)
```

+ Look carefully at the elements of $values and $vectors. What kind of numbers are these?
  Dig in with the typeof() function to figure out their type.
```{r}
typeof(z$values)
typeof(z$vectors)
```

+ if have set your code up properly, you should be able to re-run it and create a matrix of different size because n_dims will change.
YES! Its working :)

<br />

##### Exercise 2  
<br />

Create a list with the following named elements:  

+ my_matrix, which is a 4 x 4 matrix filled with random uniform values
```{r}
my_matrix <- matrix(data=runif(16), nrow=4, ncol=4)
print(my_matrix)
```

+ my_logical which is a 100-element vector of TRUE or FALSE values. Do this efficiently by setting up a vector of random values and then applying an inequality to it.
```{r}
a <- runif(100)
my_logical <- a>0.6
print(my_logical)
```

+ my_letters, which is a 26-element vector of all the lower-case letters in random order.
```{r}
my_letters <- c(sample(letters, size = 26, replace = FALSE))
print(my_letters)
```

+ now creating a list with all of these named elements
```{r}
big_list <- list(my_matrix,
                 my_logical,
                 my_letters)
print(big_list)
```

+ create a new list, which has the element[2,2] from the matrix, the second element of the logical vector, and the second element of the letters vector.
```{r}
my_list <- list(my_matrix[2,2],
                my_logical[2],
                my_letters[2])
print(my_list)

```

+ use the typeof() function to confirm the underlying data types of each component in this list
```{r}
typeof(my_list[[1]])
typeof(my_list[[2]])
typeof(my_list[[3]])
```

+ combine the underlying elements from the new list into a single atomic vector with the c() function.
```{r}
my_vector <- c(my_list[[1]], my_list[[2]], my_list[[3]])
print(my_vector)
```

+ what is the data type of this vector?
```{r}
typeof(my_vector)
```

<br />

##### Exercise 3
<br />

Create a data frame with two variables (= columns) and 26 cases (= rows).  
+ Call the first variable my_unis and fill it with a random uniform value from 0 to 10
+ Call the second variable my_letters and fill it with 26 capital letters in random order.
```{r}
my_unis <- runif(26, min = 0, max = 10)
my_letters <- c(sample(LETTERS, size = 26, replace = FALSE))
my_df <- data.frame(my_unis, my_letters, stringsAsFactors = FALSE)
head(my_df)
```

+ for the first variable, use a single line of code in R to select 4 random rows and replace the numerical values in those rows with NA.
```{r}
library(naniar)
my_df2 <- my_df %>% replace_with_na(replace = list(my_unis = sample(my_unis, size = 4, replace = FALSE)))

# a simpler way that doesn't require the library - how Lauren did it
my_df[sample(length(my_unis), size = 4, replace = FALSE),1] <- NA

```

+ for the first variable, write a single line of R code to identify which rows have the missing values.
```{r}
# is.na(my_df$my_unis)  gives me a boolean vector, assigning TRUE to the NA slots. 
# But this is not an easy way to view where the NAs are

# To get the number of rows that have NAs, its better to use the following
which(!complete.cases(my_df$my_unis)) # find NA slots

```


+ for the second variable, sort it in alphabetical order
```{r}
my_df <- my_df[order(my_df$my_letters),]
head(my_df)
```


+ calculate the column mean for the first variable.
```{r}
mean(my_df$my_unis, na.rm = TRUE)
```

